Linting Linearity in Core/System FC

### 2023-01-12 report.tex
Introduction on linear types

@@

As an example, consider the following linearly typed program which allocates
some memory and frees it twice. The variable $p$ is a linear resource created by
the \textbf{malloc} keyword.
\begin{code}
let p = malloc 4
 in free p
    free p
\end{code}

Because $p$ is a linear variable, it must be used exactly once. However, we are
using it in two different calls to \textbf{free}. The example program
\emph{doesn't typecheck}! In section~\ref{linear-types} we describe the
fundamentals of linear types in more detail with a formal definition and many
examples.

@@

Linearity in Haskell stands out from linearity in Rust and Idris2 for, at least,
the two following distinct reasons:

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception. It is an especially difficult endeavour to
        add linear types to a well-established language with a large library
        ecosystem and many active users, rather than to develop the language
        from the ground up with linear types in mind, and the successful
        approach as implemented in GHC 9.0, the leading Haskell compiler, was
        based on Linear Haskell\cite{linearhaskell}, where a linear type system
        designed with retaining backwards-compatibility and allowing code reuse
        across linear and non-linear users of the same libraries in mind was
        described. We describe Linear Haskell in detail in
        section~\ref{linear-haskell-section}.
    \item Linear types permeate Haskell down to (its) \textbf{Core}, the intermediate
        language Haskell is desugared to. \textbf{Core} is a very small
        explicitly typed language to which multiple Core-to-Core optimizing
        transformations are applied in succession. The final Core program can be
        typechecked very fast (because of the explicit type annotations) and
        doing so serves as a sanity check to the correction of the
        source transformations. If the resulting optimized Core program doesn't
        typecheck, something \emph{very wrong} has happened in a transformation!
        We discuss Core in detail in section~\ref{core-section}.
    % TODO: \item values in rust are linear by default while non-linear is the
        % haskell default?
\end{itemize}

@@

However, currently, typechecking linearity in Core is broken in the face of
optimizations! The reason is not evidently clear: if we can typecheck linearity
in the surface level Haskell why can't we in Core? In fact, if we were to
typecheck the programs in Core as they were at the surface level, we would
succeed, but programs are \emph{transformed} (by semantic-preserving optimizing
transformations) and hence end up different from how they were originally
written. For example, the following recursive let definition, where $x$ is a
linear variable that must be used exactly once, would not typecheck in a source
Haskell program since the typechecker cannot tell that $x$ is used linearly, but
this program might occur naturally in Core from transformations on a program
that did succesfully typecheck:
\begin{code}
letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

@@

A linear type system is a type system in which linear resources must be used
exactly once...!

% A linear type system is a type system in which statements such as "this function
% will use its argument exactly once" can be formally expressed and checked.

Girard's \emph{linear logic} suggests a new \emph{type system} for functional
languages through the Curry-Howard isomorphism -- every linear logic proposition
maps to a different linear type.

Values belonging to a linear type must be used exactly once. For example, the
type corresponding to the linear implication ($\lolli$), the linear function, is
a function that consumes its argument exactly once. A linear function that
receives an integer as an input and consumes it exactly once to
produce a character would have type $\textbf{int} \lolli
\textbf{char}$.

Programming languages with linear type systems might differ in their definition of linear
resources, but usually \emph{variables} are considered linear resources?
% That is, we can rely on the type system to prove properties regarding the usage of certain resources.
% The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
% In a non-memory managed language, an allocated pointer must be freed
% \emph{exactly once}.

% \begin{code}
% let p = malloc(5)
%     p' = store(p, "x")
%     (p'', v) = load(p)
%     free p''
% \end{code}

% We'd like to hide the boilerplate involved \cite{Linear constraints}

\mypara{Type Systems} A type system can be formally described through
a set of inference rules that inductively define a judgment of the
form $\Gamma \vdash M : A$, stating that program expression $M$ has
type $A$ according to the \emph{typing assumptions} for variables
tracked in $\Gamma$. For instance,
$x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
$y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
with a given type $A$ if one can construct a typing derivation with $M :
A$ as its conclusion, by repeated application of the inference rules.

The simply-typed $\lambda$-calculus is a typed core functional
language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
functional abstraction, written $\lambda x{:}A.M$, denoting
an (anonymous) function that takes an argument of type $A$, bound to
$x$ in $M$; and application $M\,N$, with the standard meaning, and
variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
taking an argument of type $A$ and returning it back.

\mypara{Linear Logic}

Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
resource-aware logic, where propositions are interpreted as resources
that are consumed during the inference process.  Where in standard
propositional logic we are able to use an assumption as many times as
we want, in linear logic every resource (i.e., every assumption) must
be used \emph{exactly once}, or \emph{linearly}. This usage
restriction gives rise to new logical connectives, based on the way
the ambient resources are used. For instance, conjunction, usually
written as $A\wedge B$, appears in two forms in linear logic:
multiplicative or simultaneous conjunction (written $A\tensor B$); and
additive or alternative conjunction (written $A\with
B$). Multiplicative conjunction denotes the simultaneous availability
of resources $A$ and $B$, requiring both of them to be
used. Alternative conjunction denotes the availability of $A$ and $B$,
but where only one of the two resources may be used. Similarly,
implication becomes linear implication, written $A\lolli B$,
denoting a resource that will consume (exactly one) resource $A$ to
produce a resource $B$.

To present the formalization of this logic, besides the new
connectives, we need to introduce the \emph{resource-aware context}
$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
also a list of variables and their types, but where each and every
variable must be used exactly once during inference.  So, to introduce
the connective $\tensor$ which defines a multiplicative pair of
propositions, we must use exactly all the resources
($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$)
proposition $A$, and ($\Delta_2$) proposition $B$:
\[
    \infer*[right=($\tensor I$)]
    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
\]
Out of the logical connectives, we need to mention one more, since it
augments the form of the judgment and it's the one that ensures
logical strength i.e. we're able to translate intuitionistic logic
into linear logic.  The proposition $\bang A$ (read \emph{of course}
$A$) is used (under certain conditions) to make a resource
``infinite'' i.e. to make it useable an arbitrary number of times. To
distinguish the ``infinite'' variables, a separate, unrestricted,
context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
resources, and $\Delta$ the resources that can only be used once.  The
linear typing judgment for the introduction of the exponential $\bang A$
takes the form:
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \emptyset \vdash M : A}
    {\Gamma ; \emptyset \vdash \bang M : \bang A}
\]
Logically, a proof of $\bang A$ cannot use linear resources since
$\bang A$ denotes an unbounded (potentially $0$) number of copies of
$A$. Proofs of $\bang A$ may use other unrestricted or exponential
resources, tracked by context $\Gamma$.
From a computational perspective, the type $\bang A$
internalizes the simply-typed $\lambda$-calculus in the linear
$\lambda$-calculus.

The elimination form for the exponential, written $\llet{ !u
= M}{ N}$, warrants the use of resource $A$ an unbounded
number of times in $N$ via the variable $u$:
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]

% TODO: Revise the story above instead of dumping so much information

In order to give well defined semantics to a linear type system, we present a
linearly typed lambda calculus which is a very simple language with linear
types.

\[
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
       \dots
   \end{array}
\]

% \mypara{Linear Types}
\subsection{Linear Types}

Girard's \emph{linear logic} suggests a new \emph{type system} for functional
languages through the Curry-Howard isomorphism -- every linear logic proposition
maps to a different linear type.

Values belonging to a linear type must be used exactly once. For example, the
type corresponding to the linear implication ($\lolli$), the linear function, is
a function that consumes its argument exactly once. A linear function that
receives an integer as an input and consumes it exactly once to
produce a character would have type $\textbf{int} \lolli
\textbf{char}$.

@@

\section{Work in progress}

\subsection{In Result of TcGblEnv axioms}

I don't understand this one yet. I would guess it has to do with multiplicity
coercions. Incorrect incompatible branch: CoAxBranch (src/Prelude/Linear/GenericUtil.hs:112:3-51):
\begin{code}
type family Fixup (f :: Type -> Type) (g :: Type -> Type) :: Type -> Type where
  Fixup (D1 c f) (D1 _c g) = D1 c (Fixup f g)
  Fixup (C1 c f) (C1 _c g) = C1 c (Fixup f g)
  Fixup (S1 c f) (S1 _c (MP1 m f)) = S1 c (MP1 m f) -- error in this constructor
  Fixup (S1 c f) (S1 _c f) = S1 c f
  Fixup (f :*: g) (f' :*: g') = Fixup f f' :*: Fixup g g'
  Fixup (f :+: g) (f' :+: g') = Fixup f f' :+: Fixup g g'
  Fixup V1 V1 = V1
  Fixup _ _ = TypeError ('Text "FixupMetaData: representations do not match.")
\end{code}

% \section{Core to Core Validation}
\subsection{Inlining}

\subsection{CSE}

\subsection{Join Points}

When duplicating a case (in the case-of-case transformation), to avoid code
explosion, the branches of the case are first made into join points

\begin{code}
case e of
  Pat1 -> u
  Pat2 -> v
~~>
let j1 = u in
let j2 = v in
case e of
  Pat1 -> j1
  Pat2 -> j2
\end{code}

If there is any linear variable in u and v, then the standard
let rule above will fail (since j1 occurs only in one branch, and
so does j2).

However, if j1 and j2 were annotated with their usage environment,

\subsection{Empty Case}

For case expressions, the usage environment is computed by checking all branches
and taking sup. However, this trick doesn't work when there are no branches.

\begin{itemize}
\item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
\item https://gitlab.haskell.org/ghc/ghc/-/issues/18768

\item (1) Just like a case expression remembers its type (Note [Why does Case have a
'Type' field?] in Core.hs), it should remember its usage environment. This data
should be verified by Lint.

\item (2) Once this is done, we can remove the Bottom usage and the second field of
UsageEnv. In this step, we have to infer the correct usage environment for empty
case in the typechecker.
\end{itemize}

\begin{code}
{-# LANGUAGE LinearTypes, EmptyCase #-}
module M where

{-# NOINLINE f #-}
f :: a %1-> ()
f x = case () of {}
\end{code}

This example is well typed: a function is linear if it consumes its argument
exactly once when it's consumed exactly once. It seems like the function isn't
linear since it won't consume x because of the empty case, however, that also
means f won't be consumed due to the same empty case, thus linearity is
preserved.

\begin{code}
* In the case of empty types (see Note [Bottoming expressions]), say
  data T
we do NOT want to replace
  case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
because x might raise an exception, and *that*'s what we want to see!
(#6067 is an example.) To preserve semantics we'd have to say
   x `seq` error Bool "Inaccessible case"
but the 'seq' is just such a case, so we are back to square 1.
\end{code}

There are three different problems:

\begin{itemize}
\item castBottomExpr converts (case x :: T of {}) :: T to x.
\item Worker/wrapper moves the empty case to a separate binding
\item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
    case in GHC.CoreToStg.Prep
\end{itemize}

With castBottomExpr, we get the example above to
\begin{code}
    f = \ @a (x (%1) :: a) -> ()
\end{code}
And if we don't, 
\begin{code}
    f = \ @a (x (%1) :: a) -> case () of {}
\end{code}
And that supposedly if we had a usage environment in the case expression we
could avoid the error. How is it typed without the transformation in face
of the bottom? (Even knowing that theoretically it's because of divergence?)

>>>>

Static type systems have brought programmers ...
Programming languages rely on type systems to statically eliminate classes of
errors ...

@@

As an example, consider the following C-like program in which allocated memory
is freed twice. We know this to be the dreaded double-free error which will
crash our programs and crush our dreams. Regardless, a C-like type system will
accept this program without any issue.
\begin{code}
let p = malloc 4;
 in free p;
    free p;
\end{code}

Under the lens of a linear type system, consider the variable $p$ to be a linear
resource created by the \textbf{malloc} keyword. Because $p$ is a linear
variable, it must be used exactly once. However, we are using it in two
different calls to \textbf{free}. With our new linearity assumptions, the
% TODO:
example program \emph{would not typecheck}! Some conclusion on linearity
goodness. In section~\ref{linear-types} we describe the fundamentals of linear
types in more detail with a formal definition and many examples.

@@

Linearity in Haskell stands out % from linearity in Rust and Idris2
for the two following distinct reasons:

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception, unlike Rust and Idris2 which were
        designed with linearity from the start. It is an especially difficult
        endeavour to add linear types to a well-established language with a
        large library ecosystem and many active users, rather than to develop
        the language from the ground up with linear types in mind, and the
        successful approach as implemented in GHC 9.0, the leading Haskell
        compiler, was based on Linear Haskell\cite{linearhaskell}, where a
        linear type system designed with retaining backwards-compatibility and
        allowing code reuse across linear and non-linear users of the same
        libraries in mind was described. We describe Linear Haskell in detail in
        section~\ref{linear-haskell-section}.

    \item Linear types permeate Haskell down to (its) \textbf{Core}, the
        intermediate language Haskell is desugared to. \textbf{Core} is a very
        small explicitly typed functional language to which multiple
        Core-to-Core optimizing transformations are applied in succession. The
        final Core program can be typechecked very fast (because of the explicit
        type annotations) and doing so serves as a sanity check to the
        correction of the source transformations. If the resulting optimized
        Core program doesn't typecheck, something \emph{very wrong} has happened
        in a transformation!  We discuss Core in detail in
        section~\ref{core-section}.
        % TODO: \item values in rust are linear by default while non-linear is
        % the haskell default?
\end{itemize}

@@

However, currently, programs transformed by optimizations can fail to typecheck
in Core because of the newly added linearity! The reason is not evidently clear:
if we can typecheck linearity in the surface level Haskell why can't we in Core?
In fact, if we were to typecheck the programs in Core as they were at the
surface level, we would succeed, but programs are \emph{transformed} (by
semantic-preserving optimizing transformations) and hence end up different from
how they were originally written. For example, the following recursive let
definition, where $x$ is a linear variable that must be used exactly once, would
not typecheck in a source Haskell program since the typechecker cannot tell that
$x$ is used linearly, but this program might occur naturally in Core from
transformations on a program that did succesfully typecheck:
\begin{code}
letrec f = \case
        True -> f False
        False -> x
in f True
\end{code}

@@

Much the same way type systems can statically eliminate various kinds of
programs that would fail at runtime, such as a program that derreferences an
integer value rather than a pointer, linear type systems can guarantee that
certain errors (regarding resource usage) are forbidden.

In linear type systems, so called linear resources must be used \emph{exactly
once}. Not using a linear resource at all or using said resource multiple times
will result in a type error. We can model many real-world resources such as file
handles, socket descriptors, and allocated memory, as linear resources. This
way, because a file handle must be used exactly once, forgetting to close the
file handle is a type error, and closing the handle twice is also a type error.
With linear types, avoiding leaks and double frees is no longer a programmer's
worry because the compiler can guarantee the resource is used exactly once, or
\emph{linearly}.

To understand how linear types are defined and used in practice, we present two
examples of anonymous functions, try to understand how they could be disregarded
as incorrect, and work our way up to linear types from them. The first receives a
file handle as an argument, ignores it, and returns $\star$ (read unit), which
is equivalent to C's
\textbf{void}.
% TODO: Better code formatting
\[
    \lambda h.~\textrm{return}~\star;
\]

And, similarly, the second receives a file handle and closes it
twice. Note that \textbf{close} has type $\textbf{Handle} \to \star$, that is, it's a
function that receives a handle and returns unit.
\[
    \lambda h.~\textrm{close}~h;~\textrm{close}~h;
\]

Both the above functions share the same type ($\textbf{Handle} \to \star$),
under a simple type system such as C's they are type correct, and both have
erroneous behaviours. The first forgets to close the handle and the second
closes it twice. Our goal is now to reach a type system that rejects these
two programs.

The key observation is that the argument of the function must be linear, and we
can augment the common function type to account for the linearity of its
argument. We take the function type $A \to B$ and replace the arrow with the
$\lolli$
%
\footnote{Since linear types are born from a
correspondence\cite{curry-howard} (the Curry-Howard isomrphism) with linear
logic\cite{linear-logic}, we borrow the $\lolli$ symbol and other linear logic
connectives to describe linear types.}
%
operator to denote a function that uses its argument exactly once: $A \lolli B$.
% Linear types are a powerful idea because they allow us to statically reason
% about resources in our program. A so called linear resource

% For example, to define a ...

% Linear types are a powerful idea

In order to give well defined semantics to a linear type system, we present a
linearly typed lambda calculus which is a very simple language with linear
types...

\[
  \begin{array}{lcl}
    s,t & ::= & \star\\
        & \mid & t \lolli s\\
        & \mid & t \oplus s\\
        & \mid & t \tensor s\\
        & \mid & t \with s\\
        & \mid & \bang t\\
   \end{array}
\]

% A linear type system is a type system in which linear resources must be used
% exactly once...!

% % A linear type system is a type system in which statements such as "this function
% % will use its argument exactly once" can be formally expressed and checked.

% Girard's \emph{linear logic} suggests a new \emph{type system} for functional
% languages through the Curry-Howard isomorphism -- every linear logic proposition
% maps to a different linear type.

% Values belonging to a linear type must be used exactly once. For example, the
% type corresponding to the linear implication ($\lolli$), the linear function, is
% a function that consumes its argument exactly once. A linear function that
% receives an integer as an input and consumes it exactly once to
% produce a character would have type $\textbf{int} \lolli
% \textbf{char}$.

% Programming languages with linear type systems might differ in their definition of linear
% resources, but usually \emph{variables} are considered linear resources?
% % That is, we can rely on the type system to prove properties regarding the usage of certain resources.
% % The \emph{Hello World} of linearly typed programs is a linearly typed pointer.
% % In a non-memory managed language, an allocated pointer must be freed
% % \emph{exactly once}.

% % \begin{code}
% % let p = malloc(5)
% %     p' = store(p, "x")
% %     (p'', v) = load(p)
% %     free p''
% % \end{code}

% % We'd like to hide the boilerplate involved \cite{Linear constraints}

% \mypara{Type Systems} A type system can be formally described through
% a set of inference rules that inductively define a judgment of the
% form $\Gamma \vdash M : A$, stating that program expression $M$ has
% type $A$ according to the \emph{typing assumptions} for variables
% tracked in $\Gamma$. For instance,
% $x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
% that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
% $y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
% with a given type $A$ if one can construct a typing derivation with $M :
% A$ as its conclusion, by repeated application of the inference rules.

% The simply-typed $\lambda$-calculus is a typed core functional
% language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
% functional abstraction, written $\lambda x{:}A.M$, denoting
% an (anonymous) function that takes an argument of type $A$, bound to
% $x$ in $M$; and application $M\,N$, with the standard meaning, and
% variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
% taking an argument of type $A$ and returning it back.

% \mypara{Linear Logic}

% Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
% resource-aware logic, where propositions are interpreted as resources
% that are consumed during the inference process.  Where in standard
% propositional logic we are able to use an assumption as many times as
% we want, in linear logic every resource (i.e., every assumption) must
% be used \emph{exactly once}, or \emph{linearly}. This usage
% restriction gives rise to new logical connectives, based on the way
% the ambient resources are used. For instance, conjunction, usually
% written as $A\wedge B$, appears in two forms in linear logic:
% multiplicative or simultaneous conjunction (written $A\tensor B$); and
% additive or alternative conjunction (written $A\with
% B$). Multiplicative conjunction denotes the simultaneous availability
% of resources $A$ and $B$, requiring both of them to be
% used. Alternative conjunction denotes the availability of $A$ and $B$,
% but where only one of the two resources may be used. Similarly,
% implication becomes linear implication, written $A\lolli B$,
% denoting a resource that will consume (exactly one) resource $A$ to
% produce a resource $B$.

% To present the formalization of this logic, besides the new
% connectives, we need to introduce the \emph{resource-aware context}
% $\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
% also a list of variables and their types, but where each and every
% variable must be used exactly once during inference.  So, to introduce
% the connective $\tensor$ which defines a multiplicative pair of
% propositions, we must use exactly all the resources
% ($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$)
% proposition $A$, and ($\Delta_2$) proposition $B$:
% \[
%     \infer*[right=($\tensor I$)]
%     {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
%     {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
% \]
% Out of the logical connectives, we need to mention one more, since it
% augments the form of the judgment and it's the one that ensures
% logical strength i.e. we're able to translate intuitionistic logic
% into linear logic.  The proposition $\bang A$ (read \emph{of course}
% $A$) is used (under certain conditions) to make a resource
% ``infinite'' i.e. to make it useable an arbitrary number of times. To
% distinguish the ``infinite'' variables, a separate, unrestricted,
% context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
% resources, and $\Delta$ the resources that can only be used once.  The
% linear typing judgment for the introduction of the exponential $\bang A$
% takes the form:
% \[
%     \infer*[right=($\bang I$)]
%     {\Gamma ; \emptyset \vdash M : A}
%     {\Gamma ; \emptyset \vdash \bang M : \bang A}
% \]
% Logically, a proof of $\bang A$ cannot use linear resources since
% $\bang A$ denotes an unbounded (potentially $0$) number of copies of
% $A$. Proofs of $\bang A$ may use other unrestricted or exponential
% resources, tracked by context $\Gamma$.
% From a computational perspective, the type $\bang A$
% internalizes the simply-typed $\lambda$-calculus in the linear
% $\lambda$-calculus.

% The elimination form for the exponential, written $\llet{ !u
% = M}{ N}$, warrants the use of resource $A$ an unbounded
% number of times in $N$ via the variable $u$:
% \[
%     \infer*[right=($\bang E$)]
%     {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
%     {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
% \]

% % TODO: Revise the story above instead of dumping so much information

% \mypara{Linear Types}
% \subsection{Linear Types}

% Girard's \emph{linear logic} suggests a new \emph{type system} for functional
% languages through the Curry-Howard isomorphism -- every linear logic proposition
% maps to a different linear type.

% Values belonging to a linear type must be used exactly once. For example, the
% type corresponding to the linear implication ($\lolli$), the linear function, is
% a function that consumes its argument exactly once. A linear function that
% receives an integer as an input and consumes it exactly once to
% produce a character would have type $\textbf{int} \lolli
% \textbf{char}$.

@@

% \section{Work in progress}

% % \section{Core to Core Validation}
% \subsection{Inlining}

% \subsection{CSE}

% \subsection{Join Points}

% When duplicating a case (in the case-of-case transformation), to avoid code
% explosion, the branches of the case are first made into join points

% \begin{code}
% case e of
%   Pat1 -> u
%   Pat2 -> v
% ~~>
% let j1 = u in
% let j2 = v in
% case e of
%   Pat1 -> j1
%   Pat2 -> j2
% \end{code}

% If there is any linear variable in u and v, then the standard
% let rule above will fail (since j1 occurs only in one branch, and
% so does j2).

% However, if j1 and j2 were annotated with their usage environment,

% \subsection{Empty Case}

% For case expressions, the usage environment is computed by checking all branches
% and taking sup. However, this trick doesn't work when there are no branches.

% \begin{itemize}
% \item https://gitlab.haskell.org/ghc/ghc/-/issues/20058
% \item https://gitlab.haskell.org/ghc/ghc/-/issues/18768

% \item (1) Just like a case expression remembers its type (Note [Why does Case have a
% 'Type' field?] in Core.hs), it should remember its usage environment. This data
% should be verified by Lint.

% \item (2) Once this is done, we can remove the Bottom usage and the second field of
% UsageEnv. In this step, we have to infer the correct usage environment for empty
% case in the typechecker.
% \end{itemize}

% \begin{code}
% {-# LANGUAGE LinearTypes, EmptyCase #-}
% module M where

% {-# NOINLINE f #-}
% f :: a %1-> ()
% f x = case () of {}
% \end{code}

% This example is well typed: a function is linear if it consumes its argument
% exactly once when it's consumed exactly once. It seems like the function isn't
% linear since it won't consume x because of the empty case, however, that also
% means f won't be consumed due to the same empty case, thus linearity is
% preserved.

% \begin{code}
% * In the case of empty types (see Note [Bottoming expressions]), say
%   data T
% we do NOT want to replace
%   case (x::T) of Bool {}   -->   error Bool "Inaccessible case"
% because x might raise an exception, and *that*'s what we want to see!
% (#6067 is an example.) To preserve semantics we'd have to say
%    x `seq` error Bool "Inaccessible case"
% but the 'seq' is just such a case, so we are back to square 1.
% \end{code}

% There are three different problems:

% \begin{itemize}
% \item castBottomExpr converts (case x :: T of {}) :: T to x.
% \item Worker/wrapper moves the empty case to a separate binding
% \item CorePrep eliminates empty case, just like point 1 (See -- Eliminate empty
%     case in GHC.CoreToStg.Prep
% \end{itemize}

% With castBottomExpr, we get the example above to
% \begin{code}
%     f = \ @a (x (%1) :: a) -> ()
% \end{code}
% And if we don't, 
% \begin{code}
%     f = \ @a (x (%1) :: a) -> case () of {}
% \end{code}
% And that supposedly if we had a usage environment in the case expression we
% could avoid the error. How is it typed without the transformation in face
% of the bottom? (Even knowing that theoretically it's because of divergence?)
